Union and intersection contracts are hard, actually

### 2021-06-10 paper.tex
Conclusion

The  design and implementation of programming languages is a challenging task. It
is a delicate exercise to balance complexity, expressivity, efficiency, and user
experience all together. Unfortunately for designers, a language cannot usually
be reduced to a simple sum of small independent features. They are rather
composed of a lot of moving parts that interact in unexpected ways. The more a
language grows, the more any new addition has to be weighted carefully, as it
has the potential to interact with any of the existing features.

Part of recent developments revolves around improving safety of languages,
ideally with little to none effort required by the programmers.  Contracts are
crucial to this endeavour, they are fundamental to allow a safe interaction
between typed and untyped code in gradually typed languages; but they are also
interesting in themselves in a fully dynamically typed language (and even in a
fully statically typed language). Contracts are a precious idiom for data
validation or to enforce runtime pre- and post-condition that would be hard to
ensure statically.

% In spirit, contracts are similar to
% assertions, but they are part of a principled framework, they are composable,
% they have good error reporting and they are integrated as a language feature. In
% particular, contracts natively handles higher-order functions and built-in data
% structures. This makes them an great candidate for integration in the Nickel
% configuration language.

% This is only the beginning of the story. A language designer then has to choose
% what contract combinators to provide among a number of possibilities: there is
% the standard package including higher-order function contracts, record contracts
% and list contracts.  Then, one could add polymorphic
% contracts~\cite{BlameForAll}, dependent function
% contracts~\cite{DependentContracts}, and so on. For validation purpose, union and
% intersection contracts are really appealing. Unions have a intuitive
% interpretation as modeling alternatives and thus appear implicitly or explicitly
% in a lot of data structures.

When trying to enforce properties on dynamic languages, unions appear naturally,
since they allow to express that a variable is in an unknown state, but within
some bounded set of alternatives.  Sadly, the implementation of union and
intersection contracts is not as simple as it first appears. In conjunction with
higher-order contracts, they introduce side-effects in the language. These
behaviors were, for most of them, already known: we do not claim their
discovery. However, as language implementers, we have made an actual attempt at
including them in a concrete language. We exposed what, in our opinion, are the
most important barriers and their impact on the rest of the language. Even in a
pure setting, union and intersection contracts break the equivalence between
call-by-name and call-by-value, and can consequently prevent optimizations.
There are discrepancies between the expected and actual semantics that make some
behaviors unintuitive. Their implementation is a non-trivial change that
contaminates the evaluation of all other contracts, and the performance penalty
is not obvious to assess.

To conclude, let us make clear that we do not think union and intersection
contracts are fundamentally broken, that they can not be implemented correctly,
or that they do not bear any value (quite the contrary). They may still make
sense to have in a language, and some apparent difficulties in the
implementation could be lifted some day. But as often, there are gaps between
the theoretical foundation, a proof-of-concept, a prototype, and the integration
in an actual language. We hope that our attempt may serve as a cautionary tale:
for union and intersection contracts, these gaps might be quite large in places.

>>>>

Despite the fact that union and intersection of dynamic properties may
at first appear like an easy task, as soon as they are combined with
higher-order contracts for increased accuracy of error messages, they
really aren't.

The problem of union and intersection contracts is that they are not
orthogonal to apparently independent features of programming
languages. The mere presence of intersection and union contract
induces computational effects, and can make it quite difficult to
perform simple program optimizations such as inlining.

Designing a language with union and intersection contracts means
making difficult choices: some features of union and intersection
contracts and of the rest of the language must be abandoned. Various
trade-offs can be made, but it is worth mentioning that implementing a
system with union and intersection contracts appears to be pretty
complex a task when unions and intersections are fairly complete.

It's hard not to have sympathy for the minimalist end of this
spectrum, where, like Racket, a language only has a very simple notion
of unions and intersections. Many applications of unions and
intersections are not possible in such a context, but the presence of
union and intersection contracts doesn't interact with the rest of the
language. It's probably more manageable.

To conclude, let us make clear that we do not think union and intersection
contracts are fundamentally broken, that they can not be implemented correctly,
or that they do not bear any value (quite the contrary). They may still make
sense to have in a language, and some apparent difficulties in the
implementation could be lifted some day. But as often, there are gaps between
the theoretical foundation, a proof-of-concept, a prototype, and the integration
in an actual language. We hope that our attempt may serve as a cautionary tale:
for union and intersection contracts, these gaps may be larger than
they appear.
